---
main_topsize: 0.13 #percent coverage of the poster
main_bottomsize: 0.1
#ESSENTIALS
author:
  - name: '**Roshan Joji**'
    affil: 1
    main: true
    email: 'roshan.joji@students.plymouth.ac.uk'
  - name: '**Kirin Gola**'
    affil: 1
    main: true
    email: 'kirin.gola@students.plymouth.ac.uk'
  - name: '**Nikhil Roy Sebastian**'
    affil: 1
    main: true
    email: 'nikhil.roysebastian@postgrad.plymouth.ac.uk'
  - name: '**Jacob Amoah**'
    affil: 1
    main: true
    email: 'jacob.amoah@students.plymouth.ac.uk'
    
affiliation:
  - num: 1
    address: Poster Group D, School of Engineering, Computing and Mathematics, University of Plymouth
main_findings:
  - "**The Role of GPUs in Modern Data Science: Performance, Cost, and Computational Demand**"
logoleft_name: secamlogo.png
logoright_name: sande.png
logocenter_name: 
output: 
  posterdown::posterdown_betterport:
    self_contained: false
    pandoc_args: --mathjax
    number_sections: false
bibliography: packages.bib
link-citations: true
knit: pagedown::chrome_print
---

```{r setup, include=FALSE}
library(tidyverse)
library(rpart)
library(randomForest)
library(class)
library(ggplot2)
```

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html") 
```

# Introduction

- Heavy computational power is needed to handle large datasets and train complex models.
- GPUs are better suited for parallel processing than CPUs, making them very effective for training neural networks, image recognition, and large-scale data analysis [1]. 
- Such high-performance hardware can be expensive and energy-intensive, raising sustainability and equity issues in the field [2].
- This poster aims to show how computational demands in machine learning (ML) scale with data size, and how GPU performance, cost, and power characteristics affect modern data science workloads.

```{r, include=FALSE}
knitr::write_bib(c('posterdown', 'rmarkdown','pagedown'), 'packages.bib')
```

# Computational Cost of Machine Learning Models

- To investigate computational cost of ML models as dataset size increases, three models were evaluated on synthetic data: Decision Tree, Random Forest, and K-Nearest Neighbours (KNN).
- Models were trained using 30 repeated timed runs for each dataset size.
- Time was recorded and averaged to provide a measure of computational cost.

```{r computation cost data synthesis and model training, include=FALSE}
# Decision Tree
results <- data.frame()

sizes <- c(1000, 10000, 20000, 30000)
reps <- 30

for (n in sizes) {
  
  for (i in 1:reps) {
    
    X <- data.frame(matrix(rnorm(n * 50), nrow = n))
    y <- sample(c("A", "B"), n, replace = TRUE)
    df <- data.frame(X, y = as.factor(y))
    
    t1 <- system.time({
      model_dt <- rpart(y ~ ., data = df)
    })
    
    results <- rbind(results,
                     data.frame(Model = "Decision Tree", Size = n, Time = t1[3]))
  }
}


results_summary <- results %>%
  group_by(Model, Size) %>%
  summarise(Time = mean(Time))

results

ggplot(results_summary, aes(x = Size, y = Time, colour = Model)) +
  geom_line() +
  geom_point()

# KNN
results_knn <- data.frame()

sizes <- c(1000, 10000, 20000, 30000)
reps <- 30

for (n in sizes) {
  
  for (i in 1:reps) {
    
    X_knn <- data.frame(matrix(rnorm(n * 50), nrow = n))
    y_knn <- sample(c("A", "B"), n, replace = TRUE)
    
    t_knn <- system.time({
      pred_knn <- knn(train = X_knn, test = X_knn, cl = y_knn, k = 5)
    })
    
    results_knn <- rbind(results_knn,
                         data.frame(Model = "KNN", Size = n, Time = t_knn[3]))
  }
}

 results_knn_summary <- results_knn %>%
  group_by(Model, Size) %>%
  summarise(Time = mean(Time))

results_knn_summary

ggplot(results_knn_summary, aes(x = Size, y = Time, colour = Model)) +
  geom_line() +
  geom_point()

# Random Forest
results_rf <- data.frame()

sizes <- c(1000, 10000, 20000, 30000)
reps <- 30

for (n in sizes) {
  
  for (i in 1:reps) {
    
    X_rf <- data.frame(matrix(rnorm(n * 50), nrow = n))
    y_rf <- sample(c("A", "B"), n, replace = TRUE)
    df_rf <- data.frame(X_rf, y = as.factor(y_rf))
    
    t_rf <- system.time({
      model_rf <- randomForest(y ~ ., data = df_rf, ntree = 100)
    })
    
    results_rf <- rbind(results_rf,
                        data.frame(Model = "Random Forest", Size = n, Time = t_rf[3]))
  }
}

results_rf_summary <- results_rf %>%
  group_by(Model, Size) %>%
  summarise(Time = mean(Time))

results_rf_summary

ggplot(results_rf_summary, aes(x = Size, y = Time, colour = Model)) +
  geom_line() +
  geom_point()

# Combine all results
results_all <- rbind(results_summary, results_knn_summary, results_rf_summary)
```

```{r, fig.width=8, fig.height=6.5, echo=FALSE, fig.cap="Logarithmic graph of training time for three machine learning models on datasets of increasing magnitude"}
ggplot(results_all, aes(x = Size, y = Time, colour = Model)) +
  geom_line(size = 1.6) +
  geom_point(size = 3) +
  scale_y_log10() +
  labs(
    x = "Dataset Size",
    y = "Time log10 (s)",
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 18, face = "bold"),
    axis.title.x = element_text(margin = ggplot2::margin(t = 15)),
    axis.title.y = element_text(margin = ggplot2::margin(r = 15)),
    axis.text = element_text(size = 16),
    legend.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 14),
    plot.background = element_rect(fill = "#00808010", colour = "white", linewidth = 0.5)
  )
```

- Computational cost generally increases with dataset size for all models, although they each scale differently.
- Minor fluctuations exist (e.g. Decision Tree) but general trend is upwards.
- As a result, more powerful hardware is required to cope with larger and more complex datasets.
- GPUs can provide the capability for parallel computations and allow for better performance when training larger, more complex models.

# GPU Performance Over Time

- GPU benchmark scores from 2010-2022 were analysed to observe how computational capability has evolved over time [3].

```{r, fig.width=8, fig.height=6.5, echo=FALSE, fig.cap="Evolution of GPU performance from 2010 - 2022 (Dots represent raw benchmark score data)"}
gpu <- read_csv("GPU_benchmarks_v7.csv")

# Create summary separately
gpu_summary <- gpu %>%
  filter(!is.na(testDate), testDate >= 2010) %>%
  group_by(testDate) %>%
  summarise(mean_score = mean(G3Dmark, na.rm = TRUE), .groups = "drop")

p3 <- ggplot() +
  
  # Raw data
  geom_point(
    data = gpu %>% filter(!is.na(testDate), testDate >= 2010),
    aes(x = testDate, y = G3Dmark),
    colour = "#008080",
    alpha = 0.4,
    size = 1.5,
    position = position_jitter(width = 0.1)
  ) +
  
  # Mean line
  geom_line(
    data = gpu_summary,
    aes(x = testDate, y = mean_score),
    size = 1.6,
    colour = "#7a2cbf"
  ) +
  
  # Mean points
  geom_point(
    data = gpu_summary,
    aes(x = testDate, y = mean_score),
    size = 3,
    colour = "#ff7b00"
  ) +
  
  labs(
    title = "             ",
    x = "Year",
    y = "Average G3DMark Score"
  ) +
  
  scale_x_continuous(breaks = seq(2010, 2022, by = 2)) +
  
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 18, face = "bold"),
    axis.title.x = element_text(margin = ggplot2::margin(t = 15)),
    axis.title.y = element_text(margin = ggplot2::margin(r = 15)),
    axis.text = element_text(size = 16),
    plot.background = element_rect(fill = "#00808010", colour = "white", linewidth = 0.5)
  )

p3
```

- GPU performance has steadily increased, with peak benchmark scores in 2022 being approx. 7 times higher than in 2010.
- Modern GPUs are far more capable at large-scale and complex modelling than in the past, allowing them to support the increasing computational demand of modern data science, and assisting in its development.

# Price vs. Performance

- To evaluate how GPU performance scales with cost, benchmark scores were compared against market value [3].

```{r, fig.width=8, fig.height=6.5, echo=FALSE, fig.cap="Relationship between GPU price and performance"}
p2 <- gpu %>%
  filter(price > 0, price < 3000) %>%
  ggplot(aes(x = price, y = G3Dmark)) +
  geom_point(alpha = 0.7, colour = "#0d58c7", size = 3) +
  geom_smooth(method = "lm", se = FALSE, colour = "red", linewidth = 1.6) +
  labs(
    title = "                     ",
    x = "Price ($)",
    y = "G3DMark Score"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size =18, face =  "bold"),
    axis.title.x = element_text(margin = ggplot2::margin(t = 15)),
    axis.title.y = element_text(margin = ggplot2::margin(r = 15)),
    axis.text = element_text(size = 16),
    plot.background = element_rect(fill = "#00808010", colour = "white", linewidth = 0.5)
  )

p2
```

- GPU performance generally increases with price.
- While the peak of GPU performance is found at the highest price points (≈ 2000 USD), excellent baseline capability is found at lower price levels as well (≈ 800-1000 USD), suggesting that cost does not solely determine performance.
- Efficient hardware selection can enable strong data science performance without requiring the most expensive GPUs.

# Power vs. Performance

- To evaluate how GPU performance scales with energy usage, thermal design power (TDP) was compared against benchmark scores [3].

```{r, fig.width=8, fig.height=6, echo=FALSE, fig.cap="Relationship between GPU TDP and performance"}
p4 <- gpu %>%
  filter(TDP > 0) %>%
  ggplot(aes(x = TDP, y = G3Dmark)) +
  geom_point(alpha = 0.7, colour = "#7a2cbf", size = 3) +
  geom_smooth(method = "lm", se = FALSE, colour = "#ff7b00", linewidth = 1.6) +
  labs(
    title = "                   ",
    x = "TDP (Watts)",
    y = "G3DMark Score"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title = element_text(size = 18, face = "bold"),
    axis.title.x = element_text(margin = ggplot2::margin(t = 15)),
    axis.title.y = element_text(margin = ggplot2::margin(r = 15)),
    axis.text = element_text(size = 16),
    plot.background = element_rect(fill = "#00808010", colour = "white", linewidth = 0.5)
  )

p4
```

- GPU performance generally increases with power consumption. However, there is substantial variation in performance at similar power usage levels.
- While peak performance is found at higher power usage, high performance GPUs can be found at moderate TDP levels.
- Again, efficient hardware selection can allow for effective data science performance without requiring excessive power consumption, thereby reducing costs associated with such.

# Conclusion

- GPUs support modern data science by providing the required processing capabilities needed for complex datasets and models.
- GPU performance has improved significantly over time, supporting the development and scalability of data science.
- While peak performance often comes with steep prices, high performance GPUs can be found at lower price points, reducing the issue of equity within the field.
- The same can be said for energy consumption, as efficient hardware selection can provide high performance with lower energy requirements, addressing sustainability concerns.

# References
1. Pitta K, Chang S. Basics on NVIDIA GPU Hardware Architecture - HECC Knowledge Base. Nasa.gov. 2025 [cited 2026 Apr 23]. Available from: https://www.nas.nasa.gov/hecc/support/kb/basics-on-nvidia-gpu-hardware-architecture_704.html2.
2. Harris D. Sustainable Strides: How AI and Accelerated Computing Are Driving Energy Efficiency. NVIDIA Blog. 2024 [cited 2026 Apr 23]. Available from: https://blogs.nvidia.com/blog/accelerated-ai-energy-efficiency/
3. GPU Benchmarks Compilation. www.kaggle.com. 2022 [cited 2026 Apr 28]. Available from: https://www.kaggle.com/datasets/alanjo/gpu-benchmarks